---
title: "Afterschool Programs’ Effect on Poverty-Adjusted Performance"
date: "April 12th, 2016"
output: pdf_document
---

```{r, echo=FALSE, message=FALSE}
library(reshape2)
library(ggplot2)
```

```{r echo=FALSE}
# Read into a variable from csv, keeping only columns we care about
poverty <- read.csv("../data/FreeReducedLunch.csv", stringsAsFactors = FALSE)
poverty <- poverty[, c(1, 5, 8, 11, 14, 17)]
# Eliminate incomplete cases
poverty <- poverty[complete.cases(poverty), ]
# head(poverty)
```

```{r, echo=FALSE}
# Read into a variable from csv, keeping only columns we care about
performance <- read.csv("../data/Performance.csv", stringsAsFactors = FALSE)
performance <- performance[, c(3, 6, 7, 9, 10)]
# Shorten terms, column names
performance$Subject[performance$Subject == "English Language Arts"] <- "Eng"
performance$Subject[performance$Subject == "Mathematics"] <- "Math"
performance$Subject[performance$Subject == "Science"] <- "Sci"
performance$Subgroup[performance$Subgroup == "American Indian/Alaskan Native"] <- "NaAmer"
performance$Subgroup[performance$Subgroup == "Black/African American"] <- "Black"
performance$Subgroup[performance$Subgroup == "Econo. Disadvantaged"] <- "EcoDis"
performance$Subgroup[performance$Subgroup == "Extended Assessment"] <- "ExAsmt"
performance$Subgroup[performance$Subgroup == "Hispanic/Latino"] <- "HisLat"
performance$Subgroup[performance$Subgroup == "Indian Education"] <- "IndEd"
performance$Subgroup[performance$Subgroup == "Limited English Proficient (LEP)"] <- "LimEng"
performance$Subgroup[performance$Subgroup == "Migrant Education"] <- "MigEdu"
performance$Subgroup[performance$Subgroup == "Multi-Racial"] <- "Multi"
performance$Subgroup[performance$Subgroup == "Pacific Islander"] <- "PacIsl"
performance$Subgroup[performance$Subgroup == "Students with Disabilities (SWD)"] <- "SWD" 
performance$Subgroup[performance$Subgroup == "SWD with Accommodations"] <- "SWDAcc" 
performance$Subgroup[performance$Subgroup == "Talented and Gifted (TAG)"] <- "TAG"
performance$Subgroup[performance$Subgroup == "Total Population"] <- "Total"
names(performance)[4:5] <- c("Part", "Met")
# Clean up data, cast as numeric
performance$Part <- sub("[<>] ([0-9]{1,2}.[0-9])%", "\\1", performance$Part)
performance$Part[performance$Part %in% c("*", "-")] <- NA
performance$Part <- as.numeric(performance$Part)
performance$Met <- sub("[<>] ([0-9]{1,2}.[0-9])%", "\\1", performance$Met)
performance$Met[performance$Met %in% c("*", "-")] <- NA
performance$Met <- as.numeric(performance$Met)
# Reshape into one row per school, one column per Group_Subject_Metric combo
performance <- recast(performance, SchoolID ~ Subgroup + Subject + variable, id.var = 1:3)
# subset to overall data, schools that have compete data
overall <- performance[, c(1, 100:105)]
overall <- overall[complete.cases(overall), ]
# merge in poverty data via percent eligible for free/reduced lunches
overall <- merge(poverty[, c(1:2, 6)], overall)
names(overall) <- sub("Total_", "", names(overall))# add school type
types <- read.csv("../data/RCmediaSchoolsAggregate.csv")[, c(4,8)]
overall <- merge(types, overall)
overall$SchoolType <- factor(overall$SchoolType, c("E", "M", "H"))
```

```{r echo=FALSE}
# fit new linear models, dependent on school type
fit_eng <- lm(Eng_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
overall$ResEng <- overall$Eng_Met - predict(fit_eng, overall)
fit_math <- lm(Math_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
overall$ResMath <- overall$Math_Met - predict(fit_math, overall)
fit_sci <- lm(Sci_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
overall$ResSci <- overall$Sci_Met - predict(fit_sci, overall)
```

```{r, echo=FALSE}
programs <- read.csv("../data/OASK_DB.csv")
programs <- programs[!is.na(programs$SchoolID), 1]
programs <- as.data.frame(table(programs))
names(programs) <- c("SchoolID", "Programs")
overall <- merge(overall, programs, all.x = TRUE)
overall$Programs[is.na(overall$Programs)] <- 0
overall$Programs[overall$Programs >= 1] <- "1+"
overall$Programs <- as.factor(overall$Programs)
# table(overall$Programs)
```

```{r echo=FALSE}
# aggregate means
agg_eng <- aggregate(ResEng ~ SchoolType + Programs, data = overall, FUN = mean)
agg_math <- aggregate(ResMath ~ SchoolType + Programs, data = overall, FUN = mean)
agg_sci <- aggregate(ResSci ~ SchoolType + Programs, data = overall, FUN = mean)
agg <- merge(agg_eng, agg_math)
agg <- merge(agg, agg_sci)
# Prepare for graphing
agg <- melt(agg, id.vars = 1:2, variable.name = "Subject")
levels(agg$SchoolType) = c("Elementary", "Middle", "High")
levels(agg$Subject) <- c("English", "Math", "Science")
```

## Abstract

Afterschool programs can serve low income kids, working to lower the achievement gap and level the playing field.  These programs are chronically under-funded in Oregon; low-income programs are often publicly funded and are unable to sustain themselves without this stream of funding.

We look at the effect of summer and afterschool programs on the performance of Oregon schools after controlling for the effects of poverty.  Our conclusion is that programming has a strongly positive correlation with performance, but with some caveats about the data we have available.

## Summary

We looked at how well 993 schools did in each subject as compared to schools of a similar poverty level and grade level, and computed averages.  The results are summarized below.

```{r echo=FALSE}
g <- ggplot(data = agg, aes(x = Programs, y = value, group = 1))
g <- g + geom_point(aes(color = SchoolType))
g <- g + facet_grid(. ~ Subject)
g <- g + ylim(-10, 10)
g <- g + ylab("Performance Relative to Expectations (%)")
g <- g + xlab("Number of Programs")
g <- g + ggtitle("Performance by School Type and Subject, 2014-15")
g
```

There are several caveats that must be acknowledged along with this analysis:

 * We do not have complete data on the majority of the over 5000 schools in Oregon
 * We have much more data on elementary schools than on middle or high schools
 * Most of the schools with complete data also host programs

## Methodology

First, we describe the data analyzed in this report.  We start by combining enrollment, free or reduced lunch eligibility (FRLE), and standardized test performance on the schools for which complete data exists.  Using FRLE rates as a metric of poverty, we first model the effect of poverty on school performance so that we may focus instead on the effect of afterschool programs.

In the following charts, we show the percent of students that met or exceeded the standard in a subject (English, Math, Science) as a function of the FRLE rate.  The size of each dot reflects the school's enrollment, and the color reflects its grade level.  We have also added lines of best fit.

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Eng_Met))
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment, color = SchoolType))
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
p <- p + geom_smooth(method = lm, se = FALSE, aes(color = SchoolType))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded English Proficiency (%)")
p <- p + ggtitle("English Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Math_Met))
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment, color = SchoolType))
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
p <- p + geom_smooth(method = lm, se = FALSE, aes(color = SchoolType))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded Math Proficiency (%)")
p <- p + ggtitle("Math Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Sci_Met))
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment, color = SchoolType))
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
p <- p + geom_smooth(method = lm, se = FALSE, aes(color = SchoolType))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded Science Proficiency (%)")
p <- p + ggtitle("Science Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

To focus on the effect of afterschool programs, we compute the difference between a school's percentage of students that meet or exceed the standard in each subject and the expected percentage based on FRLE rate.  The result is a residual performance rating for each school, where a rating of 3 would suggest that the school exceeded its expected performance in a subject by 3 percentage points.

```{r echo=FALSE}
p <- ggplot(overall, aes(x = SchoolType, y = ResEng))
p <- p + geom_boxplot(aes(fill = Programs))
p <- p + scale_fill_manual(values = c("red", "green"))
p <- p + xlab("Grade Level")
p <- p + ylab("Residual English Proficiency (%)")
p <- p + ggtitle("Residual English Proficiency v Programs Held, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = SchoolType, y = ResMath))
p <- p + geom_boxplot(aes(fill = Programs))
p <- p + scale_fill_manual(values = c("red", "green"))
p <- p + xlab("Grade Level")
p <- p + ylab("Residual Math Proficiency (%)")
p <- p + ggtitle("Residual Math Proficiency v Programs Held, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = SchoolType, y = ResSci))
p <- p + geom_boxplot(aes(fill = Programs))
p <- p + scale_fill_manual(values = c("red", "green"))
p <- p + xlab("Grade Level")
p <- p + ylab("Residual Science Proficiency (%)")
p <- p + ggtitle("Residual Science Proficiency v Programs Held, 2014-15")
p
```