---
title: "Summer Programs’ Effect on Poverty-Adjusted Performance"
author: "Kyle Joecken"
date: "March 21, 2016"
output: pdf_document
---

```{r, echo=FALSE, message=FALSE}
library(reshape2)
library(ggplot2)
```

## Abstract

We look at the effect of summer and food programs on the performance of Oregon schools after controlling for the effects of poverty.

## Controlling for Poverty

First, we need to load the data into data frames and clean what we have.  We'll read poverty data in first, dropping rows of schools on which we don't have complete data (many of these are Jails/Juvenile facilities or EI/ECSE programs):

```{r}
# Read into a variable from csv, keeping only columns we care about
poverty <- read.csv("../data/FreeReducedLunch.csv", stringsAsFactors = FALSE)
poverty <- poverty[, c(1, 5, 8, 11, 14, 17)]
# Eliminate incomplete cases
poverty <- poverty[complete.cases(poverty), ]
head(poverty)
```

\newpage

We next load performance data on English and Mathematics.  These data will need to be heavily massaged to get them into a format we can proceed with.

```{r}
# Read into a variable from csv, keeping only columns we care about
performance <- read.csv("../data/Performance.csv", stringsAsFactors = FALSE)
performance <- performance[, c(3, 6, 7, 9, 10)]
# Shorten terms, column names
performance$Subject[performance$Subject == "English Language Arts"] <- "Eng"
performance$Subject[performance$Subject == "Mathematics"] <- "Math"
performance$Subject[performance$Subject == "Science"] <- "Sci"
performance$Subgroup[performance$Subgroup == "American Indian/Alaskan Native"] <- "NaAmer"
performance$Subgroup[performance$Subgroup == "Black/African American"] <- "Black"
performance$Subgroup[performance$Subgroup == "Econo. Disadvantaged"] <- "EcoDis"
performance$Subgroup[performance$Subgroup == "Extended Assessment"] <- "ExAsmt"
performance$Subgroup[performance$Subgroup == "Hispanic/Latino"] <- "HisLat"
performance$Subgroup[performance$Subgroup == "Indian Education"] <- "IndEd"
performance$Subgroup[performance$Subgroup == "Limited English Proficient (LEP)"] <- "LimEng"
performance$Subgroup[performance$Subgroup == "Migrant Education"] <- "MigEdu"
performance$Subgroup[performance$Subgroup == "Multi-Racial"] <- "Multi"
performance$Subgroup[performance$Subgroup == "Pacific Islander"] <- "PacIsl"
performance$Subgroup[performance$Subgroup == "Students with Disabilities (SWD)"] <- "SWD" 
performance$Subgroup[performance$Subgroup == "SWD with Accommodations"] <- "SWDAcc" 
performance$Subgroup[performance$Subgroup == "Talented and Gifted (TAG)"] <- "TAG"
performance$Subgroup[performance$Subgroup == "Total Population"] <- "Total"
names(performance)[4:5] <- c("Part", "Met")
# Clean up data, cast as numeric
performance$Part <- sub("[<>] ([0-9]{1,2}.[0-9])%", "\\1", performance$Part)
performance$Part[performance$Part %in% c("*", "-")] <- NA
performance$Part <- as.numeric(performance$Part)
performance$Met <- sub("[<>] ([0-9]{1,2}.[0-9])%", "\\1", performance$Met)
performance$Met[performance$Met %in% c("*", "-")] <- NA
performance$Met <- as.numeric(performance$Met)
# Reshape into one row per school, one column per Group_Subject_Metric combo
performance <- recast(performance, SchoolID ~ Subgroup + Subject + variable, id.var = 1:3)
# subset to overall data, schools that have compete data
overall <- performance[, c(1, 100:105)]
overall <- overall[complete.cases(overall), ]
# merge in poverty data via percent eligible for free/reduced lunches
overall <- merge(poverty[, c(1:2, 6)], overall)
names(overall) <- sub("Total_", "", names(overall))
```

\newpage

Let's have a look at the data as collected and see if a pattern emerges.

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Eng_Met))
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment))
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
p <- p + geom_smooth(method = lm, se = FALSE, aes(weight = Enrollment))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded English Proficiency (%)")
p <- p + ggtitle("English Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Math_Met))
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment))
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
p <- p + geom_smooth(method = lm, se = FALSE, aes(weight = Enrollment))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded Math Proficiency (%)")
p <- p + ggtitle("Math Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Sci_Met))
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment))
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
p <- p + geom_smooth(method = lm, se = FALSE, aes(weight = Enrollment))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded Science Proficiency (%)")
p <- p + ggtitle("Science Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

\newpage

In order to remove the effect that our measure of poverty has on proficiency, we simply subtract the expected proficiency rates from the reported ones to get a *residual* proficiency rate (or a performance rate relative to expectation).

```{r}
fit_eng <- lm(Eng_Met ~ Enrollment + PercentEligible, data = overall, weights = Enrollment)
overall$ResEng <- overall$Eng_Met - predict(fit_eng, overall)
fit_math <- lm(Math_Met ~ Enrollment + PercentEligible, data = overall, weights = Enrollment)
overall$ResMath <- overall$Math_Met - predict(fit_math, overall)
fit_sci <- lm(Sci_Met ~ Enrollment + PercentEligible, data = overall, weights = Enrollment)
overall$ResSci <- overall$Sci_Met - predict(fit_sci, overall)
```

Now we collect information on which schools have programs and which do not.

```{r}
programs <- read.csv("../data/OASK_DB.csv")
programs <- programs[!is.na(programs$SchoolID), 1]
programs <- as.data.frame(table(programs))
names(programs) <- c("SchoolID", "Programs")
overall <- merge(overall, programs, all.x = TRUE)
overall$Programs[is.na(overall$Programs)] <- 0
overall$Programs[overall$Programs >= 1] <- "1+"
overall$Programs <- as.factor(overall$Programs)
table(overall$Programs)
```

Finally, we plot residual school performance against the number of programs held at the school to see if a pattern emerges.

```{r echo=FALSE}
p <- ggplot(overall, aes(x = Programs, y = ResMath))
p <- p + geom_boxplot()
p <- p + xlab("Programs Held")
p <- p + ylab("Residual Math Proficiency (%)")
p <- p + ggtitle("Residual Math Proficiency v Programs Held, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = Programs, y = ResEng))
p <- p + geom_boxplot()
p <- p + xlab("Programs Held")
p <- p + ylab("Residual English Proficiency (%)")
p <- p + ggtitle("Residual English Proficiency v Programs Held, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = Programs, y = ResSci))
p <- p + geom_boxplot()
p <- p + xlab("Programs Held")
p <- p + ylab("Residual Science Proficiency (%)")
p <- p + ggtitle("Residual Science Proficiency v Programs Held, 2014-15")
p
```

# A look at mean performance by school type, subject

```{r echo=FALSE}
# drop old residual data
overall <- overall[, c(1:9, 13)]
# add school type
types <- read.csv("../data/RCmediaSchoolsAggregate.csv")[, c(4,8)]
overall <- merge(types, overall)
overall$SchoolType <- factor(overall$SchoolType, c("E", "M", "H"))
# fit new linear models, dependent on school type
fit_eng <- lm(Eng_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
overall$ResEng <- overall$Eng_Met - predict(fit_eng, overall)
fit_math <- lm(Math_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
overall$ResMath <- overall$Math_Met - predict(fit_math, overall)
fit_sci <- lm(Sci_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
overall$ResSci <- overall$Sci_Met - predict(fit_sci, overall)
# aggregate means
agg_eng <- aggregate(ResEng ~ SchoolType + Programs, data = overall, FUN = mean)
agg_math <- aggregate(ResMath ~ SchoolType + Programs, data = overall, FUN = mean)
agg_sci <- aggregate(ResSci ~ SchoolType + Programs, data = overall, FUN = mean)
agg <- merge(agg_eng, agg_math)
agg <- merge(agg, agg_sci)
# Prepare for graphing
agg <- melt(agg, id.vars = 1:2, variable.name = "Subject")
levels(agg$SchoolType) = c("Elementary", "Middle", "High")
levels(agg$Subject) <- c("English", "Math", "Science")
```

```{r echo=FALSE}
g <- ggplot(data = agg, aes(x = Programs, y = value))
g <- g + geom_bar(stat = "identity", position = "dodge")
g <- g + facet_grid(Subject ~ SchoolType)
g <- g + ylab("Performance Relative to Poverty Level (%)")
g <- g + xlab("Number of Programs")
g <- g + ggtitle("Performance by School Type and Subject, 2014-15")
g
```