---
title: "Afterschool Programs’ Effect on Poverty-Adjusted Performance"
date: "April 12th, 2016"
output: pdf_document
---

```{r, echo=FALSE, message=FALSE}
# import packages for additional functionality
library(reshape2)
library(ggplot2)
```

```{r echo=FALSE}
# Read into a variable from csv, keeping only columns we care about
poverty <- read.csv("../data/FreeReducedLunch.csv", stringsAsFactors = FALSE)
poverty <- poverty[, c(1, 5, 8, 11, 14, 17)]
# Eliminate incomplete cases
poverty <- poverty[complete.cases(poverty), ]
# head(poverty)
```

```{r, echo=FALSE}
# Read into a variable from csv, keeping only columns we care about
performance <- read.csv("../data/Performance.csv", stringsAsFactors = FALSE)
performance <- performance[, c(3, 6, 7, 9, 10)]
# Shorten terms, column names
performance$Subject[performance$Subject == "English Language Arts"] <- "Eng"
performance$Subject[performance$Subject == "Mathematics"] <- "Math"
performance$Subject[performance$Subject == "Science"] <- "Sci"
performance$Subgroup[performance$Subgroup == "American Indian/Alaskan Native"] <- "NaAmer"
performance$Subgroup[performance$Subgroup == "Black/African American"] <- "Black"
performance$Subgroup[performance$Subgroup == "Econo. Disadvantaged"] <- "EcoDis"
performance$Subgroup[performance$Subgroup == "Extended Assessment"] <- "ExAsmt"
performance$Subgroup[performance$Subgroup == "Hispanic/Latino"] <- "HisLat"
performance$Subgroup[performance$Subgroup == "Indian Education"] <- "IndEd"
performance$Subgroup[performance$Subgroup == "Limited English Proficient (LEP)"] <- "LimEng"
performance$Subgroup[performance$Subgroup == "Migrant Education"] <- "MigEdu"
performance$Subgroup[performance$Subgroup == "Multi-Racial"] <- "Multi"
performance$Subgroup[performance$Subgroup == "Pacific Islander"] <- "PacIsl"
performance$Subgroup[performance$Subgroup == "Students with Disabilities (SWD)"] <- "SWD" 
performance$Subgroup[performance$Subgroup == "SWD with Accommodations"] <- "SWDAcc" 
performance$Subgroup[performance$Subgroup == "Talented and Gifted (TAG)"] <- "TAG"
performance$Subgroup[performance$Subgroup == "Total Population"] <- "Total"
names(performance)[4:5] <- c("Part", "Met")
# Find instances of '> 95.0%' or '< 5.0%' and replace them with '95.0' and '5.0'
performance$Part <- sub("[<>] ([0-9]{1,2}.[0-9])%", "\\1", performance$Part)
performance$Met <- sub("[<>] ([0-9]{1,2}.[0-9])%", "\\1", performance$Met)
# Replace '*' and '-' with NA
performance$Part[performance$Part %in% c("*", "-")] <- NA
performance$Met[performance$Met %in% c("*", "-")] <- NA
# Cast proportions as numeric, so that we can do math on them (they are read in as strings)
performance$Part <- as.numeric(performance$Part)
performance$Met <- as.numeric(performance$Met)
# Reshape into one row per school, one column per Group_Subject_Metric combo (reshape2)
performance <- recast(performance, SchoolID ~ Subgroup + Subject + variable, id.var = 1:3)
# Subset to overall data, schools that have compete data
overall <- performance[, c(1, 100:105)]
overall <- overall[complete.cases(overall), ]
# Merge in poverty data via percent eligible for free/reduced lunches
overall <- merge(poverty[, c(1:2, 6)], overall)
# Remove the 'Total_' before the metrics in the column names
names(overall) <- sub("Total_", "", names(overall))
# Read in and add school type
types <- read.csv("../data/RCmediaSchoolsAggregate.csv")[, c(4,8)]
overall <- merge(types, overall)
# Reorder the SchoolType factor so that they appear in a natural order
overall$SchoolType <- factor(overall$SchoolType, c("E", "M", "H"))
```

```{r echo=FALSE}
# Fit linear models, dependent on school type
fit_eng <- lm(Eng_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
fit_math <- lm(Math_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
fit_sci <- lm(Sci_Met ~ PercentEligible * SchoolType, data = overall, weights = Enrollment)
# Compute how far removed each school's performances are relative to
# the expectations of the above models (residual performance)
overall$ResEng <- overall$Eng_Met - predict(fit_eng, overall)
overall$ResMath <- overall$Math_Met - predict(fit_math, overall)
overall$ResSci <- overall$Sci_Met - predict(fit_sci, overall)
```

```{r, echo=FALSE}
# Read in the programs list
programs <- read.csv("../data/OASK_DB.csv")
# Since the only thing we need for this analysis is a count of how many
# programs exist at each school, we simply count how many times each
# SchoolID appears in this list.
# This line grabs the SchoolID column in every row where SchoolID isn't NA.
programs <- programs[!is.na(programs$SchoolID), 1]
# The table function is a quick way to count what's in a vector.
programs <- as.data.frame(table(programs))
# Rename the columns
names(programs) <- c("SchoolID", "Programs")
# Add the counts to the data set we've collected
overall <- merge(overall, programs, all.x = TRUE)
# If a school didn't have a program in the file, we change the NA to a 0
overall$Programs[is.na(overall$Programs)] <- 0
# To keep things simple, we collapse all non-zero counts down to '1+'
overall$Programs[overall$Programs >= 1] <- "1+"
overall$Programs <- as.factor(overall$Programs)
# table(overall$Programs)
```

```{r echo=FALSE}
# Compute the mean residual performance in each subject, grouped by
# SchoolType and whether or not the school hosts a program
agg_eng <- aggregate(ResEng ~ SchoolType + Programs, data = overall, FUN = mean)
agg_math <- aggregate(ResMath ~ SchoolType + Programs, data = overall, FUN = mean)
agg_sci <- aggregate(ResSci ~ SchoolType + Programs, data = overall, FUN = mean)
# Put all the means back into one data frame
agg <- merge(agg_eng, agg_math)
agg <- merge(agg, agg_sci)
# Instead of one row for each subject, we collect the means
# into separate columns so that each school is a row
agg <- melt(agg, id.vars = 1:2, variable.name = "Subject")
# Human-readable labels
levels(agg$SchoolType) = c("Elementary", "Middle", "High")
levels(agg$Subject) <- c("English", "Math", "Science")
```

## Abstract

Afterschool programs can serve low income kids, working to lower the achievement gap and level the playing field.  These programs are chronically under-funded in Oregon; low-income programs are often publicly funded and are unable to sustain themselves without this stream of funding.

We look at the effects of summer and afterschool programs on the English, Math, and Science performance of Oregon schools.  After controlling for the general effects of poverty, we find that presence of programs has a strong positive correlation with school performance.

## Summary

We compared the performance of a total of 993 schools across three subjects (English, Math, Science) and three school types (Elementary, Middle, and High Schools).  The key results are summarized below.

```{r echo=FALSE}
# These commands generate a plot.  As they're all basically the same,
# we'll just go through this one and update when new methods are used.
# We first define the data set and assign variables to columns
g <- ggplot(data = agg, aes(x = Programs, y = value, group = 1))
# Next we plot a point for each class of school, colored by SchoolType
g <- g + geom_point(aes(color = SchoolType))
# We break the figure into three, one for each Subject
g <- g + facet_grid(. ~ Subject)
# Force symmetric axes
g <- g + ylim(-10, 10)
# Add x- and y-axis labels and a title
g <- g + ylab("Performance Relative to Expectations (%)")
g <- g + xlab("Number of Programs")
g <- g + ggtitle("Performance by School Type and Subject, 2014-15")
g
```

The diagram above shows that for each of the three subjects, the average performance of a given school type (e.g. Elementary) is higher for the schools that have afterschool and summer programs.

There are several caveats that must be acknowledged along with this analysis:

 * We do not have complete data on the majority of the over 5000 schools in Oregon
 * We have much more data on elementary schools than on middle or high schools
 * Most of the schools with complete data also host programs

## Methodology

First, we describe the data analyzed in this report.  We start by combining enrollment, free or reduced lunch eligibility (FRLE), and standardized test performance on the schools for which complete data exists.  Using FRLE rates as a metric of poverty, we first model the effect of poverty on school performance so that we may focus instead on the effect of afterschool programs.

In the following charts, we show the percent of students that met or exceeded the standard in a subject (English, Math, Science) as a function of the FRLE rate.  The size of each dot reflects the school's enrollment, and the color reflects its grade level.  We have also added lines of best fit.

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Eng_Met))
# Alpha affects the points' opacity; we also size each dot by Enrollment
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment, color = SchoolType))
# We want size to scale with area, so that a school twice as big look so
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
# We add lines of best fit, one for each SchoolType
p <- p + geom_smooth(method = lm, se = FALSE, aes(color = SchoolType))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded English Proficiency (%)")
p <- p + ggtitle("English Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Math_Met))
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment, color = SchoolType))
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
p <- p + geom_smooth(method = lm, se = FALSE, aes(color = SchoolType))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded Math Proficiency (%)")
p <- p + ggtitle("Math Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = PercentEligible, y = Sci_Met))
p <- p + geom_point(alpha = 0.5, aes(size = Enrollment, color = SchoolType))
p <- p + scale_size_area()
p <- p + xlim(0, 100) + ylim(0, 100)
p <- p + geom_smooth(method = lm, se = FALSE, aes(color = SchoolType))
p <- p + xlab("Students Eligible for Free/Reduced Lunches (%)")
p <- p + ylab("Met/Exceeded Science Proficiency (%)")
p <- p + ggtitle("Science Proficiency v Free/Reduced Lunch Eligibility, 2014-15")
p
```

To focus on the effect of afterschool programs, we compute the difference between a school's percentage of students that meet or exceed the standard in each subject and the expected percentage based on FRLE rate.  The result is a residual performance rating for each school, where a rating of 3 would suggest that the school exceeded its expected performance in a subject by 3 percentage points.

```{r echo=FALSE}
# To give a fuller picture of the distribution of the residual performance 
# data, we use box-and-whisker plots.
p <- ggplot(overall, aes(x = SchoolType, y = ResEng))
# Make box-and-whisker plots, coloring the boxes by existance of Programming
p <- p + geom_boxplot(aes(fill = Programs))
# Override the default color choices
p <- p + scale_fill_manual(values = c("red", "green"))
p <- p + xlab("Grade Level")
p <- p + ylab("Residual English Proficiency (%)")
p <- p + ggtitle("Residual English Proficiency v Programs Held, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = SchoolType, y = ResMath))
p <- p + geom_boxplot(aes(fill = Programs))
p <- p + scale_fill_manual(values = c("red", "green"))
p <- p + xlab("Grade Level")
p <- p + ylab("Residual Math Proficiency (%)")
p <- p + ggtitle("Residual Math Proficiency v Programs Held, 2014-15")
p
```

```{r echo=FALSE}
p <- ggplot(overall, aes(x = SchoolType, y = ResSci))
p <- p + geom_boxplot(aes(fill = Programs))
p <- p + scale_fill_manual(values = c("red", "green"))
p <- p + xlab("Grade Level")
p <- p + ylab("Residual Science Proficiency (%)")
p <- p + ggtitle("Residual Science Proficiency v Programs Held, 2014-15")
p
```